Extracting sunbeams from cucumbers: Reproducibility as a necessity for the future of archaeological science

### 2025-01-30 paper.qmd
recommendations

## Methods and materials

@@

## Discussion

@@

Similarly bleak results come from measurements specifically of the reproducibility of computational analyses of scientific studies. An attempt at reproducing the computational results of 204 papers in *Science* resulted in success in reproducing the findings for 26% [@stoddenEmpiricalAnalysisJournal2018]. The computational results of two out of 41 geoscience papers could be fully reproduced on the first attempt [@konkolComputationalReproducibilityGeoscientific2019]. In the biomedical field, code in 1,203 out of 27,271 (4%) notebooks associated with 3,467 publications could be run without errors [@samuelComputationalReproducibilityJupyter2024]. Economists have been especially active in researching computational reproducibility, with studies indicating successful reproduction of results using code and data provided by authors for 30% of 67 papers [@ChangLi2015], 14% of 203 papers [@gertler2018make], 44% of 152 papers [@herbert2021reproducibility], 30% of 419 articles [@fivsar2024reproducibility], and 28% of 168 papers [@perignonComputationalReproducibilityFinance2024]. These efforts confirm that the reproducibility of published research is widely recognised as a cornerstone of rigorous science, and work on evaluating how successful a research community is at generating reproducible results has become a distinctive and important meta-research activity in many fields.

@@

## Investigating computational reproducibility in archaeology

In 2024 the *Journal of Archaeological Science* introduced a new kind of peer review that has provided an opportunity to tackle this question about computational reproducibility in archaeology. In January 2024 I accepted the position of 'Associate Editor for Reproducibility' (AER) for JAS and conducted reproducibility reviews of submissions that mentioned programming languages such as R or Python in the methods sections. A reproducibility review examines the code and data used to generate the results presented in the paper, and attempts to run the authors' code to reproduce the their results. This new AER role is based on similar positions (i.e. 'data editor' or 'reproducibility editor') that journals in economics [@vilhuber2019report], statistics [@wrobel2024partnering], astronomy [@muench2023roles], ecology [@bolnick2022ensuring], and environmental studies [@rosenbergReproducibleResultsPolicy2021] have had, in some cases for over a decade. In 2024, three archaeology journals, in addition to JAS, added AERs to their editorial communities: *Advances in Archaeological Practice* [@marwick2024introducing, one paper reviewed], *Journal of Field Archaeology* [@farahani2024reproducibility, two papers reviewed], and *American Antiquity* [@martin2024editors, no papers reviewed].

@@

At the time of writing (Feburary 2025) I have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS. Of these, 11 have been published in JAS so far. Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate so of 63%. Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors.

it is hard to assess whether reported findings are credible.

While not yet the norm in archaeology, the systemic verification of the computational reproducibility of the results of conditionally accepted papers is now required in many leading academic journals including the *American Economic Review*, *American Economic Journal*, *Econometrica*, *Review of Economic Studies*, *Economic Journal*, *Journal of the American Statistical Association*, *American Journal of Political Science*, and the *Journal of Politics*.

@@

```{=html}
<!-- n an ideal science, scholars share a common background of established theories, facts and methods. This allows them to agree (usually after debate and further evidence) on the validity and significance of a new research finding, making it the basis for further theorizing and research. Harder sciences are hypothesised to come closer to this ideal. Moving towards “softer” fields, this consensus becomes less likely to be reached, the common background shrinks and fractures, and so data become less able to “speak for themselves” [\[](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0066938#pone.0066938-Fanelli1)

>>>>

## How to measure the hardness or softness of a science?

@@

## Reproducibility: A key measure of how scientific a field is

@@

Similarly bleak results come from measurements specifically of the reproducibility of computational analyses of scientific studies. An attempt at reproducing the computational results of 204 papers in *Science* resulted in success in reproducing the findings for 26% [@stoddenEmpiricalAnalysisJournal2018]. The computational results of two out of 41 geoscience papers could be fully reproduced on the first attempt [@konkolComputationalReproducibilityGeoscientific2019]. In the biomedical field, code in 1,203 out of 27,271 (4%) notebooks associated with 3,467 publications could be run without errors [@samuelComputationalReproducibilityJupyter2024]. Statisticians could reproduce 15% of 93 papers [@xiongStatePlayReproducibility2023]. Economists have been especially active in researching computational reproducibility, with studies indicating successful reproduction of results using code and data provided by authors for 30% of 67 papers [@ChangLi2015], 14% of 203 papers [@gertler2018make], 44% of 152 papers [@herbert2021reproducibility], 30% of 419 articles [@fivsar2024reproducibility], and 28% of 168 papers [@perignonComputationalReproducibilityFinance2024]. These efforts confirm that the reproducibility of published research is widely recognised as a cornerstone of rigorous science, and work on evaluating how successful a research community is at generating reproducible results has become a distinctive and important meta-research activity in many fields.

@@

## How reproducible is archaeology? Investigating computational reproducibility

In 2024 the *Journal of Archaeological Science* introduced a new kind of peer review that has provided an opportunity to tackle this question about computational reproducibility in archaeology. In January 2024 I accepted the position of 'Associate Editor for Reproducibility' (AER) for JAS and conducted reproducibility reviews of submissions that mentioned programming languages such as R or Python in the methods sections. A reproducibility review examines the code and data used to generate the results presented in the paper, and attempts to run the authors' code to reproduce the their results (see @editorsReproducibilityJournalArchaeological for more details about this process). This new AER role is based on similar positions (i.e. 'data editor' or 'reproducibility editor') that journals in economics [@vilhuber2019report], statistics [@wrobel2024partnering], astronomy [@muench2023roles], ecology [@bolnick2022ensuring], and environmental studies [@rosenbergReproducibleResultsPolicy2021] have had, in some cases for over a decade. In 2024, three archaeology journals, in addition to JAS, added AERs to their editorial communities: *Advances in Archaeological Practice* [@marwick2024introducing, one paper reviewed], *Journal of Field Archaeology* [@farahani2024reproducibility, two papers reviewed], and *American Antiquity* [@martin2024editors, no papers reviewed].

@@

At the time of writing (Feburary 2025) I have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS. Of these, 11 have been published in JAS to date. Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate of 63%. Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors. For comparison with reproducibility studies in other fields reported above, the seven fully reproducible papers should be divided by the 25 reviewed for reproducibility, resulting in a 28% success rate. Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97), which are articles that could have been eligible for reproducibility review had the authors used an open source programming language (e.g. instead of commercial software such as Microsoft Excel or SPSS, etc.) the success rate is 7%. In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values obtained from a variety of harder and soft sciences.

@fig-aer-summary shows a summary of basic characteristics of the 25 articles that have been through to the reproducibility review so far. The most commonly used software is R, followed by Python. Results generated with proprietary or closed-source software are out of scope for reproducibility reviews. Several distinct types of analyses are well-represented in this sample, especially geometric morphometry, network statistics, and analyses using artificial intelligence or machine learning algorithms (this includes deep learning and neural networks). Most authors are sharing their code and data files via Zenodo, a non-profit generic research data repository hosted by CERN that accepts any file format and freely assigns all publicly available uploads a DOI to make the files easily and uniquely citable [@petersZenodoSpotlightTraditional2017]. In this same category of DOI-issuing, research-grade repositories is OSF (the Open Science Foundation), Figshare, and university repositories. GitHub, a commercial service owned by Microsoft is a code hosting platform that is convenient for collaboration, is also popular among JAS authors, but is a problematic choice because does not offer DOIs or the same commitments to long-term availability of Zenodo. Some authors attached their code and data as journal article supplementary files, but this is a poor choice for long-term availability because these files are typically renamed and converted to different formats during the article production process, making it difficult or impossible for a reader to combine the code and data to reproduce the results.

Panels D and E of @fig-aer-summary summarise the common issues that resulted in irreproducible results. The most common issue was an incomplete compendium. This ranges from missing data files down to missing lines of code, and in most cases seems to be due to carelessness. Unspecified or under-specified dependencies is another common issue that prevents code from running. This refers to the software packages in addition to R or Python that an author used to do specialized analyses and visualisations, (e.g. dplyr for R or numpy for Python). If an author does not clearly specify the name and version number of the packages that they used for their analysis, it can be very time-consuming or impossible to correctly identify these because many packages have functions with similar names, and functions in any one package can change they way they behave as the developers update their package. Other reasons why papers failed the reproducibility review is that the paths to data files were incorrectly specified (likely a result of the author reorganising their compendium after completing their analysis, or omitting data files from the materials submitted for review), and errors returned by functions, which have diverse causes.

Poor communication about dependencies is widespread in code written by researchers in many disciplines, and a common source of irreproducible results [@trisovicLargescaleStudyResearch2022; @samuelComputationalReproducibilityJupyter2024]. In our sample issues relating to dependencies are strongly associated with the use of Python. One possible reason for this is that relative to R, Python uses more package managers, more environments, and deeper dependency chains with more complex inter-dependencies that change more rapidly [@decan2016topology; @decanEmpiricalComparisonDependency2019; @korkmazModelingImpactPython2020]. Another reason may be that there is a bigger and more established community of R users in archaeology [@schmidtToolDrivenRevolutionsArchaeological2020; @batistOpenArchaeologyOpen2024] that highly values code that is easy for others to reuse [e.g. @bilottiPointPatternAnalysis2024; @willExploringUtilityUnretouched2025].

## How to improve the computational reproducibility of archaeology?

Despite the small number of reproducibility reviews reported on here, there are patterns of common issues that point to efforts that have high potential to increase reproducibility. The problem of incomplete materials can be tackled in several ways. First, authors should use a clear and logical folder structure for their code and data. There are many simple templates to guide authors on this, e.g. @doi:10.1080/00031305.2017.1375986, @figueiredoSimpleKitUse2022, \@Cookiecutter, @cooperGuideReproducibleCode2017 and @wilsonGoodEnoughPractices2017 Second, for analyses that are not highly time-consuming, authors should re-run their code more than once, and ideally on more than one machine, before submission to confirm everything works as expected. Complex and time-consuming analyses should use pipeline or workflow management tools, e.g. GNU Make, Luigi, Snakemake, or Targets, to document the relationship of the files and folders in a machine-readable format and simplify running and re-running code [@wrattenReproducibleScalableShareable2021; @landau2021targets]. Third, authors should include in their compendium a README document that describes to a reader the folders and files contained in the project. The README file is typically the first file that a reader will look at in a compendium so it should include brief instructions to guide the user to a successful reproduction of the original results (e.g. what order to run the code files in). Fourth, authors should ensure clear, direct and obvious connections between their code and the results they present in their paper. A simple way to do this is to have one code file for each figure and table, and name the code files with the figure or table number and some key words in the caption. Another way some authors are accomplishing this is by using literate programming tools, such as Quarto and Jupyter notebooks [@quarto; @kluyver2016jupyter] that enable the research narrative and code for data analysis to be woven together in one document. Archaeologists are using these tools write complete journal articles, e.g. @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024.

Dependencies

so efforts aimed at systemic improvements of dependency handling—for example, as per [134]—have great potential to increase reproducibility.... it is hard to assess whether reported findings are credible.

While not yet the norm in archaeology, the systemic verification of the computational reproducibility of the results of conditionally accepted papers is now required in many leading academic journals including the American Economic Review, American Economic Journal, Econometrica, Review of Economic Studies, Economic Journal, Journal of the American Statistical Association, American Journal of Political Science, and the Journal of Politics.

@@

n an ideal science, scholars share a common background of established theories, facts and methods. This allows them to agree (usually after debate and further evidence) on the validity and significance of a new research finding, making it the basis for further theorizing and research. Harder sciences are hypothesised to come closer to this ideal. Moving towards “softer” fields, this consensus becomes less likely to be reached, the common background shrinks and fractures, and so data become less able to “speak for themselves” [\[](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0066938#pone.0066938-Fanelli1)
